Variability as a better characterization of Shannon entropy

### 2021-02-02 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Preparing new arxiv submission

%\documentclass[prb,preprint]{revtex4-1}

\documentclass{article}
\textheight 8.9 in \textwidth 6.5 in \oddsidemargin -10pt \topmargin -30pt

\usepackage{amsmath}
%\usepackage[usenames,dvipsnames]{color}

\title{Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula}

\author{
	Gabriele Carcassi,
	Julian Barbour and
	Christine A. Aidala 
}

@@

@@

The use of Shannon entropy $-\sum p_i \log p_i$ is well established in very different fields of knowledge such as computer science[cite], physics[cite] and ecology[cite]. Within computer science it is a fundamental element in information theory since Shannon developed his mathematical theory of communication. It has been used extensively within statistical mechanics to link the microscopic description of a system to its macroscopic thermodynamic one. It has been used in ecology to characterize biodiversity and calculate species distributions under given constraints. Entropy maximization has been shown to be a powerful tool.

While the techniques are well developed, it is puzzling that this concept can be applied in such different contexts, given that it assumes different features in each one. For example, in information theory the Shannon entropy for a book will change whether we choose to encode words or letters, which means the quantity does not have a unique value: it depends on an arbitrary choice. This is not problematic for an information theoretic quantity, but it is for a physical quantity. In statistical mechanics the $p_i$ represent probabilities, which means the Shannon entropy is a property of an ensemble of physical systems. This leads some to claim that the Shannon entropy is epistemic in nature and represents ignorance. [Add references] In ecology, however, the $p_i$ represent the fraction of individuals belonging to the $i$th species. There is no probabilistic ensemble to which one can associate a degree of ignorance. Finally, entropy in physics is associated to the notion of states and irreversible time evolution, which are not well defined concepts for the other fields. The same expression, then, seems to have different traits and it is not clear why a single concept would apply in all these cases.

To make matters worse, it is fairly typical to describe entropy with notions such as uncertainty[cite], knowledge[cite], lack of knowledge[cite], disorder[cite]. While some of these characterizations are useful in some contexts to give an intuitive picture, they do not always seem to work. A common objection to the characterization as disorder, for example, is that a uniform distribution corresponds to a homogeneous system, which one would expect to be maximally ordered; yet the distribution maximizes the Shannon entropy, which would indicate maximal disorder. Moreover, these notions remain vague in the sense that the formula is not formally derived from those concepts, and therefore it is not clear why that particular indicator would be an appropriate choice. We are left to wonder: is there really some single underlying concept, that, when applied appropriately in the different fields, reduces to the correct notion of entropy? Or is it just a big coincidence?

We propose that there is indeed a characterization of Shannon's formula that is both more general and more appropriate; that is independent of whether it is applied to sources of messages, ensembles of microstates or animal populations. We will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements}. That is, given a distribution of some elements over some quantity, the Shannon entropy represents how varied, how different from each other, the values are. In fact, it is the only continuous and linear index that measures the degree of variability within a distribution, and those requirements lead naturally to the same three conditions that Shannon set for his formula.\cite{Shannon} In other words, the concept of variability within a distribution leads in a formal way to the correct expression. In all cases where the formula is applied, in fact, we are using distributions as objects of our investigation and the variability within that distribution will be of interest. Yet, the physical meaning of those distributions can be widely different: they may represent the fractional distribution of animals according to their species; they may represent probability distributions of different outcomes; they may represent knowledge of certain facts as in Bayesian statistics. Therefore the variability itself, in the context of these different cases, will necessarily acquire a different meaning. In particular, we will see how this perspective naturally unfolds in the contexts of information theory and statistical mechanics.

As we believe this is the right characterization, we will therefore use the expression Shannon ``variability'', rather than ``entropy'', ``information'' or ``diversity index'', as it tells us exactly what the quantity describes. It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Boltzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. We will reserve the word ``entropy'' for the one dealing with physical states and irreversible processes.

This characterization also explains why the Shannon variability has such a broad range of applicability. As nature presents itself in a multitude of forms and combinations, a fundamental task in science is to recognize which objects belong together and study how their different attributes are expressed in different ways or degrees. They may be galaxies, animals or outcomes of the same process; we may study their shape, behavior or numerical quantities. But the idea of grouping them and trying to understand the source of their variation is a common endeavor in science.\footnote{In fact, determining which collection of objects is significant and what attributes are relevant is a significant part of the work.} Measuring and explaining this variability is an essential task. It is this very variety that is the inspiration for and the subject of scientific investigation. A universe without variety, where all objects were the same and remained the same, is barely conceivable and certainly it would leave no space for scientific investigation. In the words of Leibniz, the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.\footnote{MOVE TO BIBLIO Paragraph 58 of Leibniz's \emph{Mondadology}}

As a final note, we want to stress that the goal of this work is not to present some new technical result. The goal is instead to synthesize all details into a coherent picture that gives them context and a precise relationship to each other. This can help, especially the new practitioners, to avoid common misconceptions. Moreover, while some details presented here may be well known in a particular community, we find that they may be new to another. For instance, we will show an example of Huffmann coding to illustrate the meaning of the numerical value associated with the Shannon variability. This is standard practice in computer science courses, yet it is not in statistical mechanics. Therefore restating all the crucial aspects in one place is beneficial to make the connection between them readily available.

%https://www.researchgate.net/publication/259011656_A_tribute_to_Claude_Shannon_1916-2001_and_a_plea_for_more_rigorous_use_of_species_richness_species_diversity_and_the_'Shannon-Wiener'_Index

%https://web.stanford.edu/~hastie/Papers/maxent_explained.pdf

%https://www.mdpi.com/1099-4300/21/8/794/htm

%https://www.sciencedirect.com/science/article/abs/pii/S0169534714001037

%Leibniz
%We recall, though we have been unable to relocate, a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''. [REF]

%We conclude this section with a word on terminology. In mathematics, Cantor defined a set as ``a collection into a whole $M$ of definite and separate objects''. He called the objects \emph{elements}. We will also use the word `elements' to denote the things we consider but subject to the proviso that they have a distinguishing feature (or features) that allows us to say that \emph{they are of the same kind}. Thus, we consider sets of a restricted type. The point is this. To constitute a set in accordance with Cantor's definition, it is only necessary for its elements to be \emph{definite and separate} objects. This leaves total freedom as to how such objects in the universe are to be parcelled up into specific sets. For example, a set might consist of Mars and the word `go'. We think such latitude misses one of the most significant aspects of the universe, namely, that the things within it can be distinguished by attributes that are much more specific than the two required by Cantor. Our field of investigation goes beyond mere number, which washes out all attributes apart from the two, of being definite and separate, that allow us to count. They were the only attributes that Cantor needed; we need more.

@@

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. In accordance with the comments made in our introduction, they can for example represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different outcomes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interests and practices of scientists and their fields of study and therefore it is, in this sense, discretionary. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements will have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or interesting than others. For example, in taxonomy the set of all animals that have feathers may be more appropriate than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

@@

Having decided what elements to study and the level of description we are interested in, we bin them; that is, we group them based on that description, disregarding the identity of the particular element. What we will be interested in is only the relative frequency $p_i = N_i / N = \left| \{e_\alpha \in E | q(e_\alpha) = q_i \} \right| / N$ of the elements within each bin. We may do so because we are either not interested or not able to further distinguish the elements. Whatever the reason, once the choices of elements and properties are made, this relative frequency is objective. The nature of the $p_i$ will depend on the previous choices. It may represent the fraction of elements if the set is constituted by a group of objects. It may represent a probability of an outcome if the set consists of different realizations of similarly prepared systems. We will therefore call $p_i$ weights, to be clear that we make no commitment as to whether we have fractions, frequentist probabilities, Bayesian probabilities or other notions.

Once we have constructed our distribution, we will want to construct an indicator $H$ that quantifies how much variability the elements exhibit within the distribution. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $H$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), such statistical quantities are not well defined. Second, relabeling the values (e.g. switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g.~$1 \% / $mm) and this will change under unit transformation (e.g.~$1000 \% /$m). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is a source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We will discuss the problem in section \ref{sec_cont}.}

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the weights. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ to be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity additive when combining two independent distributions.

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
We note that these are exactly the requirements Shannon put forth for his expression\cite{Shannon}, and that these requirements lead to only one possible option: $H(p_i) = - \sum p_i \log(p_i)$.

The Shannon variability is a clear and important attribute of a given distribution: it quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon variability represents the variability of the elements in a distribution, transcending what the distribution physically signifies.

@@

The Shannon variability may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you didn't.
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon variability. Yet, you know more in the second case: you know you have won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary? All these questions are interesting, all these questions are insightful and the answers to all these questions are expressed through just one thing: the Shannon variability.

If the Shannon expression is a measure of variability, how is it connected to information? What is it measuring and in what units?

@@

We have seen that the Shannon variability measures the variability within a distribution but, in the end, it is a numeric quantity and we cannot expect to understand the concept if we do not understand what the number represents and in what units.

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like ``is it an animal?" would be redundant. Others, like ``is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. However, it is clear that there is always a minimum number of questions that must be asked to get to the answer. Given that there are more than two species, a single question cannot be enough. It should however be intuitively clear that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon variability quantifies: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the dsitribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies how much information is needed to transfer a value from a distribution.

Defining a set of questions means choosing an encoding as we are choosing how the information gets codified into our series of bits. To understand how this works, we can briefly review the Huffman coding,[CITE] which is the optimal algorithm for symbol-by-symbol coding with a known probability distribution. The idea is that we want all possible answers to each question to be balanced, to provide the same amount of information. The reason is that making one answer more specific (i.e.~it applies in fewer cases) means making another less specific (i.e.~it applies in more cases). In the case of binary questions, then, we ideally want the probability to answer yes or no to be 50\%.

For example, suppose the population of pets in a country is as follows:
\begin{itemize}
\item dogs 27\%
\item cats 48\%
\item fish 10\%
\item bird 8\%
\item small mammal 4\%
\item reptile 3\%
\end{itemize}
For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask ``is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask `` is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptile and bird with small mammal to form a 13/14 split. So we can ask ``is it a fish or a reptile?" If the answer is yes, the followup question would be ``is it a fish?" If the answer is no, the followup question would be ``is it a bird?" With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
\item bird 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
\item small mammal 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
\item reptile 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

Now that we understand that the Shannon variability is measured by the number of bits required to identify an element from a distribution, it is instructive to derive the same expression from different considerations. Suppose we have a sequence of $N$ elements, say pets like in the previous example. These are taken from $I$ different cases: dogs, cats, fish and so on. Suppose $N_i$ are the of elements of each type, which means $\sum_{i \in I} N_i = N$. Then, given a particular instance of $N_i$, we have $W(N_i) = \frac{N!}{\prod_{i \in I} N_i!}$ possible ways to realize that sequence, which corresponds to all possible permutations. If all permutations are equally likely, then $\log W (N_i)$ represents the number of bits needed to identify one of the sequences. Assuming we use base $b$ for the logarithm, we take $N$ to be large and, using Stirling's approximation $\ln N! \approx N \ln N - N$, we find:
\begin{equation}
\begin{aligned}
\log W(N_i) &= \log \frac{N!}{\prod_{i \in I} N_i!} \\
&= \log N! - \sum_{i \in I} \log N_i! \\
&\approx \frac{1}{\ln b} \left(N \ln N - N - \sum_{i \in I} N_i \ln N_i + \sum_{i \in I} N_i \right) \\
&= \frac{1}{\ln b} \left(\sum_{i \in I} N_i \ln N - \sum_{i \in I} N_i \ln N_i \right) \\
&= -\sum_{i \in I} N_i \left(\log N_i - \log N \right) \\
&= - N \sum_{i \in I} \frac{N_i}{N} \log \frac{N_i}{N} \\
&= - N \sum_{i \in I} p_i \log p_i \\
&= N H(p_i)
\end{aligned}
\end{equation}

>>>>

\documentclass[nofootinbib,twocolumn]{revtex4-1} 
% The line above defines the type of LaTeX document.
% Note that AJP uses the same style as Phys. Rev. B (prb).

% The % character begins a comment, which continues to the end of the line.

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures
\usepackage{hyperref}

@@

\title{Variability as a better characterization of Shannon entropy}

\author{Gabriele Carcassi}
\email{carcassi@umich.edu}
\author{Christine A. Aidala}
\email{caidala@umich.edu}
\affiliation{Department of Physics, University of Michigan, Ann Arbor, MI 48109-1040 }
% Please provide a full mailing address here.

\author{Julian Barbour}
\email{julian.barbour@physics.ox.ac.uk}
\altaffiliation[permanent address: ]{College Farm, South Newington, Banbury, Oxon, OX15 4JG UK}
\affiliation{Visiting Professor in Physics at the University of Oxford (2009-2018)}

\date{\today}

\begin{abstract}
	The Shannon entropy, one of the cornerstones of information theory, is widely used in physics, particularly in statistical mechanics. Yet its characterization and connection to physics remain vague, leaving ample room for misconceptions and misunderstanding. We will show that the Shannon entropy can be fully understood as measuring the variability of the elements within a given distribution: it characterizes how much variation can be found within a collection of objects. We will see that it is the only indicator that is continuous and linear, that it quantifies the number of yes/no questions (i.e.~bits) that are needed to identify an element within the distribution, and we will see how applying this concept to statistical mechanics in different ways leads to the Boltzmann, Gibbs and von Neumann entropies.
\end{abstract}

@@

The goal of this work is to give a comprehensive introduction to the concept of Shannon entropy, the expression $-\sum p_i \log p_i$, which is well established in such disparate fields as computer science, communication theory, physics \cite{Jaynes1957-1, Grandy} ecology and economics. The concept of entropy in physics, despite being treated extensively in the physics education literature (e.g. \cite{baierlein1994entropy, carson2002undergraduate, Santillan_2008, LeffPartI, LeffPartII, LeffPartIII, LeffPartIV, LeffPartV, Benguigui_2013, geller2014entropy, Tiwary_2020}), is still surrounded by a great deal of confusion \cite{Swendsen, Styer}. Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium \cite{jarzynski2017stochastic, goldstein2019gibbs, maes2003time}. Partly because the same word ``entropy'' is used to refer to concepts (e.g. topological entropy, thermodynamic entropy, graph entropy, R\'{e}nyi entropy, information entropy) that have different definitions and various degrees of overlap.  We will focus on the notion of Shannon entropy, by which we mean the use of $- \sum_i p_i \log p_i$ in the broadest characterization, which has received less attention in physics education. We will see that the Shannon entropy can be given a crisp and precise characterization and will explore its relation to the Gibbs and Boltzmann entropies in statistical mechanics.\footnote{We will not discuss its relation to thermodynamic entropy. Thermodynamic entropy is defined on the notions of heat and work while the Shannon entropy is defined on the notion of a distribution. The contexts are different and relating them goes beyond the scope of this work. } The same material is also presented in a series of videos \cite{videos}.

To give a crisp and intuitive meaning to the Shannon entropy we need to address two main problems. The first problem is a consequence of its success. As it is used in many fields, it is generally introduced with concepts that are specific to that field but may not be appropriate for others. As we will see, the meaning of the $p_i$ can vary significantly. The first goal, then, is to provide a characterization of the Shannon entropy that satisfies the following requirements:
\begin{itemize}
	\item it is defined in a general setting that is crisp and intuitive
	\item it is precise enough so that the formula itself can be derived
	\item it recovers the specific meaning when applied to each field.
\end{itemize}
In this paper we show that the idea of variability of the elements within a distribution, how diverse the objects are from each other within the collection, can serve as such a general concept. We will therefore use the term \emph{``Shannon variability''} for this more general idea that applies to all cases, leaving ``entropy'' to the more specific physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability and it measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. The variability of messages within a communication channel will correspond to its information entropy as used in communication theory; the variability of microstates within a macrostate will correspond to the Gibbs entropy used in statistical mechanics; the variability of the state of a single particle within a microstate will correspond to the Boltzmann entropy. This avoids the use of incomplete or imprecise characterizations (e.g. uncertainty, knowledge, lack of knowledge, disorder) that, while useful in some limited circumstances, promote confusion in others.

The second problem is that the Shannon formula presents some peculiarity when applied over continuous variables that is often the source of confusion. Namely, it is generally linked to a choice of unit, of the resolution used to describe the problem. However, the physical state spaces, in both classical and quantum mechanics, provide additional structure that make the Shannon variability coordinate-independent. This is seldom pointed out, and it is critical to form a proper connection to the Boltzmann, Gibbs and von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch on a few of the common misconceptions and show how they are resolved.

@@

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. In accordance with the comments made in the introduction, they can for example represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different outcomes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interests and practices of scientists and their fields of study and therefore it is, in this sense, discretionary. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or insightful than others. For example, in taxonomy the set of all animals that have feathers may be more appropriate than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

@@

Having decided what elements to study and the level of description in which we are interested, we bin them; that is, we group them based on that description, disregarding the identity of the particular element. What we will be interested in is only the relative frequency $p_i = N_i / N = \left| \{e_\alpha \in E | q(e_\alpha) = q_i \} \right| / N$ of the elements within each bin. We may do so because we are either not interested or not able to further distinguish the elements. Whatever the reason, once the choices of elements and properties are made, this relative frequency is objective. The nature of the $p_i$ will depend on the previous choices. It may represent the fraction of elements if the set is constituted by a group of objects. It may represent a probability of an outcome if the set consists of different realizations of similarly prepared systems. We will therefore call $p_i$ weights, to be clear that we make no commitment as to whether we have fractions, frequentist probabilities, Bayesian probabilities or other notions.

Once we have constructed our distribution, we want to construct an indicator $H$ that quantifies how much variability the elements exhibit within the distribution. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $H$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), such statistical quantities are not well defined. Second, relabeling the values (e.g.~switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g.~$1 \% / $mm) and this will change under unit transformation (e.g.~$1000 \% /$m). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is a source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We discuss the issue in section \ref{sec_cont}.}

@@

As noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g.~mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the weights. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ to be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity additive when combining two independent distributions.

To sum up, we have the following three requirements:
\begin{enumerate}
	\item $H$ depends only on $p_i$ and it does so continuously
	\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
	\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
These are the same requirements Shannon put forth for his expression\cite{Shannon}, from which he showed that the only possible choice is $H(p_i) = - \sum p_i \log(p_i)$. See also \cite{khinchin2013mathematical} for a similar derivation. This expression quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on.

@@

The Shannon variability may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
	\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won nothing.
	\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon variability. Yet, you know more in the second case: you know you won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos islands, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary?

If the Shannon expression is a measure of variability, why is it connected to information? How is variability quantified and in what units?

@@

To understand what the numerical value represents, consider this example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like, ``Is it an animal?" would be redundant. Others, like, ``Is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. However, there has to be a minimum number of questions that must be asked to get to the answer. A single question, for example, cannot be enough given that there are more than two species. It should be intuitively clear that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon variability quantifies: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the distribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies \emph{how much information is needed to transfer a value picked from a known distribution}.

Defining a set of questions means choosing an encoding as we are choosing how the information gets codified into our series of bits. To understand how this works, we can briefly review the Huffman coding,\cite{Huffman} which is the optimal algorithm for symbol-by-symbol coding with a known probability distribution. The idea is that we want all possible answers to each question to be balanced, to provide the same amount of information. The reason is that making one answer more specific (i.e.~it applies in fewer cases) means making another less specific (i.e.~it applies in more cases). In the case of binary questions, then, we ideally want the probability to answer yes or no to be 50\%.

For example, suppose the population of pets in a country is as follows:
\begin{table}[h]
	\begin{tabular}{lr}
		dogs & 27\% \\
		cats & 48\% \\
		fish & 10\% \\
		birds & 8\% \\
		small mammals & 4\% \\
		reptiles & 3\%
	\end{tabular}
\end{table}

\noindent
For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask, ``Is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask, ``Is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptiles and birds with small mammals to form a 13/12 split. So we can ask, ``Is it a fish or a reptile?". If the answer is yes, the followup question would be ``Is it a fish?". If the answer is no, the followup question would be ``Is it a bird?". With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
	\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
	\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
	\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
	\item birds 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
	\item small mammals 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
	\item reptiles 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

Now that we understand that the Shannon variability is measured by the number of bits required to identify an element from a distribution, it is instructive to derive the same expression from different considerations. Suppose we have a sequence of $N$ elements, say pets like in the previous example. These are taken from $I$ different cases: dogs, cats, fish and so on. Suppose $N_i$ are the number of elements of each type, which means $\sum_{i \in I} N_i = N$. Then, given a particular instance of $N_i$, we have $W(N_i) = \frac{N!}{\prod_{i \in I} N_i!}$ possible ways to realize that sequence, which corresponds to all possible permutations. If all permutations are equally likely, then $\log W (N_i)$ represents the number of bits needed to identify one of the sequences.

When $N$ is large, we can use Stirling's approximation $\ln N! \approx N \ln N - N$, and find $\log W(N_i) \approx - N \sum_{i \in I} \frac{N_i}{N} \log \frac{N_i}{N} = N H(p_i)$. The logarithm of the permutations is $N$ times the Shannon variability. The result should not be surprising: it simply tells us that encoding a sequence of $N$ elements is the same as encoding $N$ elements one at a time.
